---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  dev = "ragg_png"  # higher quality than base png
)
```

# enrichmet

<!-- badges: start -->
<!-- badges: end -->

**enrichmet** simplifies local pathway enrichment analysis of metabolomics data by allowing the complete workflow to be executed through a single R function call. This design eliminates repetitive steps such as data reformatting and parameter configuration, improving efficiency, reducing the risk of errors, and supporting reproducible analysis. For users who wish to run only a specific analysis or generate selected plots rather than executing the full workflow, **enrichmet** also provides dedicated modules for each individual analysis and visualisation.

**enrichmet** performs pathway over-representation analysis using Fisher's exact test (optionally restricted to a measured background), computes relative betweenness centrality for metabolites, and performs Metabolite Set Enrichment Analysis (MetSEA) via **fgsea**. The `enrichmet()` function produces three tables (S3 `data.frame` objects): pathway enrichment results, MetSEA results, and metabolite centrality. In addition, it generates eight plots (S3/S4 plot objects):

- **Pathway enrichment plot**
- **Pathway impact plot**
- **Metabolite Set Enrichment Analysis (MetSEA) plot**
- **Relative Betweenness Centrality (RBC) plot**
- **Pathway–metabolite network graph**
- **Pathway heatmap**
- **Pathway membership plot**
- **Reactome interaction network plot**

If differential analysis is run with `run_de()`, a volcano plot can also be included.

## Installation

You can install **enrichmet** as:

```{r install, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("enrichmet")
```

## Annotation data

Pathway and compound annotations are obtained at **runtime** from public APIs (KEGG REST and Reactome mappings) and cached locally with **BiocFileCache**. The package does **not** redistribute third-party database dumps (KEGG, Reactome, or LION). Equivalent `Pathway` / `Metabolites` tables can also be supplied by the user.

```{r kegg, eval = FALSE}
# Human pathway-to-metabolite map (KEGG organism code "hsa")
PathwayVsMetabolites <- fetch_kegg_pathway_metabolites(organism = "hsa")

# KEGG compound IDs and names, used when input_type = "name" or "mixed"
kegg_lookup <- fetch_kegg_compound_lookup()

# Reactome reaction mappings for the interaction network
kegg_ids <- extract_kegg_ids(PathwayVsMetabolites)
reactome_df <- fetch_kegg_reactome(kegg_ids = kegg_ids, species = "Homo sapiens")
```

## Input requirements

`enrichmet()` needs a **query set** of metabolites, supplied as either:

- `inputMetabolites`: a character vector of KEGG IDs and/or names, interpreted according to `input_type` (`"kegg"` (default), `"name"`, or `"mixed"`); or
- `da_results`: the output of `run_de()`, from which the query set is derived.

For MetSEA, or to build a measured-metabolite background, also provide summary statistics (`example_data`) with the columns `met_id`, `pval` or `padj`, and `log2fc`.

## Example

This is a basic example using human KEGG pathways:

```{r example}
# Load the enrichmet package
library(enrichmet)

# Retrieve annotations at runtime (cached with BiocFileCache)
PathwayVsMetabolites <- fetch_kegg_pathway_metabolites(organism = "hsa")
kegg_lookup <- fetch_kegg_compound_lookup()

# Example summary statistics
example_path <- get_cached_file(
  "https://zenodo.org/api/records/17819145/files/summary_stat.csv/content"
)
example_summary <- read.csv(example_path, stringsAsFactors = FALSE)

# Reactome reaction mappings
kegg_ids <- extract_kegg_ids(PathwayVsMetabolites)
reactome_df <- fetch_kegg_reactome(
  kegg_ids = kegg_ids,
  species = "Homo sapiens"
)

# Create example input metabolites (from example_summary)
inputMetabolites <- c(
  "C00031", "C00022", "C00197", "C00221", "C00631",
  "C01172", "C00074", "C00186", "C00036", "C00158"
)

# Run comprehensive enrichment analysis with ALL plot types
results <- enrichmet(
  inputMetabolites = inputMetabolites,
  PathwayVsMetabolites = PathwayVsMetabolites,
  example_data = example_summary,
  kegg_lookup = kegg_lookup,
  reactome_df = reactome_df,
  top_n = 10,
  p_value_cutoff = 1,
  analysis_type = c(
    "enrichment", "gsea", "centrality", "network",
    "heatmap", "membership", "interaction"
  ),
  network_top_n = 8,
  heatmap_top_n = 8,
  membership_top_n = 8,
  min_pathway_occurrence = 2,
  min_metabolite_occurrence = 1,
  input_type = "kegg"
)
```

### Results

```{r results}
cat("=== ENRICHMENT ANALYSIS RESULTS ===\n")
cat("Input metabolites used:",
    length(results$input_metabolites_used), "\n")
cat("Pathways tested:",
    nrow(results$pathway_enrichment_all), "\n")
cat("Pathways reported (p <= cutoff):",
    nrow(results$pathway_enrichment_results), "\n")

cat("\n=== TOP 5 ENRICHED PATHWAYS ===\n")
print(head(
  results$pathway_enrichment_results[
    , c("Pathway", "P_value", "Adjusted_P_value", "Enrichment_Ratio")
  ],
  5
))

cat("\n=== TOP 5 GSEA PATHWAYS ===\n")
print(head(
  results$gsea_results[
    , c("pathway", "pval", "padj", "NES")
  ],
  5
))

cat("\n=== TOP 5 CENTRAL METABOLITES ===\n")
print(head(
  results$metabolite_centrality[
    , c("Display_Name", "RBC_Metabolite")
  ],
  5
))

# Display available plots
cat("\n=== AVAILABLE PLOTS ===\n")
available_plots <- names(results)[sapply(results, function(x) {
  any(class(x) %in% c("ggplot", "gg", "ggraph", "Heatmap", "HeatmapList"))
})]
cat("Plots generated:", paste(available_plots, collapse = ", "), "\n")
```

### Visualisations

```{r plot-pathway, fig.width = 10, fig.height = 7}
# Pathway enrichment (Fisher's exact test)
results$pathway_plot
```

```{r plot-impact, fig.width = 10, fig.height = 7}
# Pathway impact: significance vs. topology-based impact score
results$impact_plot
```

```{r plot-gsea, fig.width = 10, fig.height = 7}
# MetSEA (fgsea)
results$gsea_plot
```

```{r plot-rbc, fig.width = 10, fig.height = 7}
# Relative betweenness centrality of query metabolites
results$rbc_plot
```

```{r plot-network, fig.width = 10, fig.height = 7}
# Pathway-metabolite network
results$network_plot
```

```{r plot-heatmap, fig.width = 10, fig.height = 7}
# Pathway heatmap
results$heatmap_plot
```

```{r plot-membership, fig.width = 12, fig.height = 8}
# Pathway membership
ComplexHeatmap::draw(results$membership_plot)
```

```{r plot-interaction, fig.width = 10, fig.height = 8}
# Reactome-based metabolite interaction network
results$interaction_plot
```

## Background-corrected enrichment

Fisher's exact test can be restricted to metabolites that were measured (testable) in the experiment, reducing bias from database compounds that were never observed.

```{r background, eval = FALSE}
measured_background <- extract_measured_kegg_ids(example_summary)

results_bg <- enrichmet(
  inputMetabolites = inputMetabolites,
  PathwayVsMetabolites = PathwayVsMetabolites,
  example_data = example_summary,
  kegg_lookup = kegg_lookup,
  reactome_df = reactome_df,
  backgroundMetabolites = measured_background,
  input_type = "kegg"
)
```

## Differential analysis followed by enrichment

For an end-to-end analysis that starts from the metabolomics matrix (metabolites as rows, samples as columns), run differential analysis with `run_de()` and pass the result to `enrichmet()` via `da_results`. In this case `inputMetabolites` can be left `NULL`.

```{r de-workflow, eval = FALSE}
metabolomics_path <- get_cached_file(
  "https://zenodo.org/api/records/17819145/files/example_data.csv/content"
)
metabolomics_mat <- read.csv(
  metabolomics_path,
  row.names = 1,
  check.names = FALSE
)

da_out <- run_de(
  metabolomics_mat,
  "TK-CMV",
  "K-CMV",
  fc_threshold = 1,
  pval_threshold = 0.05
)

results_da <- enrichmet(
  inputMetabolites = NULL,
  PathwayVsMetabolites = PathwayVsMetabolites,
  da_results = da_out,
  example_data = example_summary,
  kegg_lookup = kegg_lookup,
  reactome_df = reactome_df
)
```

## Other pathway maps

The same over-representation interface accepts any pathway–feature table in
`Pathway` / `Metabolites` format. Lipid ontology content is **not shipped
with the package**. Instead, LION lipid ontology content is obtained at
runtime through the BioPortal / LION programming interface when users call
the corresponding package helper, such as `fetch_lion_lipid_ontology()`.
Retrieved content is fetched on demand and cached locally (for example,
using `BiocFileCache`) to avoid unnecessary repeated downloads.

Users may also supply their own pathway–feature mapping when appropriate.
The example below uses a minimal illustrative table only.


## Data sources

- Kanehisa, M., et al. (2025). KEGG: biological systems database as a model of the real world. *Nucleic Acids Research*, 53, D672–D677. <https://www.kegg.jp/>
- Milacic, M., et al. (2024). The Reactome Pathway Knowledgebase 2024. *Nucleic Acids Research*, 52(D1), D672–D678. <https://reactome.org/>
- Molenaar, M. R., et al. (2019). LION/web: a web-based ontology enrichment tool for lipidomic data analysis. *GigaScience*, 8(6), giz061. <http://www.lipidontology.com/>
