---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# enrichmet

<!-- badges: start -->
<!-- badges: end -->

EnrichMet performs pathway enrichment analysis using Fisher’s exact test, computes betweenness centrality for metabolites,
and performs Metabolite Set Enrichment Analysis (MetSEA). It also generates plots for pathway enrichment, MetSEA, and relative betweenness centrality (RBC), and interaction plot.

## Installation

You can install enrichmet as:

``` r
install.packages("remotes")
remotes::install_git("https://github.com/biodatalab/enrichmet.git")
or 
install.packages("devtools")
devtools::install_git("https://github.com/biodatalab/enrichmet.git")
or
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("enrichmet")

```

## Example

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300,
  dev = "ragg_png"  # higher quality than base png
)
```
This is a basic example 

```{r example}
  library(igraph)
  library(ggraph)
  library(tidygraph)
  library(ggplot2)
  library(openxlsx)
  library(dplyr)
  library(tidyr)
  library(fgsea)
  library(purrr)
  library(ggrepel)
  library(stringr)
  library(pheatmap)
  library(ComplexHeatmap)
  library(ggplotify)
  library(enrichmet)
## basic example code
## ** Examples

# Generate example data with at least n=50 metabolites
set.seed(1234)

# Create 50 unique metabolites
inputMetabolites <- paste0("M", 1:20)

# ---- 1. Generate 10 pathways with random metabolites assigned ----
pathway_names <- paste0("Pathway", 1:50)
PathwayVsMetabolites <- data.frame(
  Pathway = rep(pathway_names, each = 1),
  Metabolites = sapply(1:50, function(x) paste(sample(inputMetabolites, sample(5:15, 1)), collapse = ","))
)

# ---- 2. Add new pathway entries (Pathway101 and Pathway102) ----
new_rows <- data.frame(
  Pathway = c("Pathway101", "Pathway102", "Pathway103", "Pathway104", "pathway105"),
  Metabolites = c(
    "M12,M13,M14,M15,M16,M1,M18,M3,M29,M6,M16,M4",
    "M6,M7,M8,M9,M10,M11,M9,M29,M6,M6,M16,M4",
    "M24,M25,M26,M27,M28,M29,M30,M29,M26,M5",
    "M13,M14,M15,M16,M17,M24,M27,M14",
    "M15,M16,M17,M18,M19,M20,M21,M4,M8,M10"
    
  )
)

# Combine with existing PathwayVsMetabolites
PathwayVsMetabolites <- rbind(PathwayVsMetabolites, new_rows)

# ---- 3. Generate example metabolite-level data ----
example_data <- data.frame(
  met_id = inputMetabolites,
  pval = runif(20, 0.001, 0.05),  # Random p-values between 0.001 and 0.05
  log2fc = rnorm(20, mean = 0, sd = 1)  # Log2 fold changes from normal distribution
)

# ---- 4. Create mapping_df ----
set.seed(42)
mapping_df <- data.frame(
  KEGG_ID = inputMetabolites,
  PubChem_CID = as.character(sample(10000:99999, length(inputMetabolites))),
  STITCH_ID = paste0("CIDs", str_pad(sample(1000:9999, length(inputMetabolites)), 8, pad = "0"))
)

# ---- 5. Create synthetic STITCH interaction data ----
stitch_ids <- mapping_df$STITCH_ID

stitch_pairs <- expand.grid(chemical1 = stitch_ids, chemical2 = stitch_ids) %>%
  filter(chemical1 != chemical2)

set.seed(123)
stitch_df <- stitch_pairs %>%
  slice_sample(n = 200) %>%
  mutate(
    similarity = runif(n(), 0, 1),
    experimental = sample(0:500, n(), replace = TRUE),
    database = sample(c(0, 300, 600, 900), n(), replace = TRUE),
    textmining = sample(0:1000, n(), replace = TRUE),
    combined_score = similarity * 200 + experimental + database + textmining
  ) %>%
  as_tibble()

# ---- 6. Run enrichment analysis ----
enrichmet(
  inputMetabolites = inputMetabolites,
  PathwayVsMetabolites = PathwayVsMetabolites, example_data,
  top_n = 20,
  mapping_df = mapping_df,
  stitch_df = stitch_df
)

```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300,
  dev = "png"  # or try dev = "svg" for vector graphics
)
