---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# enrichmet

<!-- badges: start -->
<!-- badges: end -->

**enrichmet** simplifies pathway enrichment analysis by allowing the complete workflow to be executed through a single R function call. This design eliminates repetitive steps such as data reformatting and parameter configuration, improving efficiency, reducing the risk of errors, and supporting reproducible analysis. For users who wish to run only a specific analysis or generate selected plots rather than executing the full workflow, enrichmet also provides dedicated modules for each individual analysis and visualization.

**enrichmet performs pathway enrichment analysis using Fisher’s exact test, computes betweenness centrality for metabolites, and performs Metabolite Set Enrichment Analysis (MetSEA). The **enrichmet**() function produces three tables (S3 data.frame objects), which may include the MetSEA table, metabolite centrality, and pathway enrichment results. In addition, it generates eight plots (S3/S4 plot objects):

- **Pathway enrichment plot**
- **Pathway impact plot**
- **Metabolite Set Enrichment Analysis (MetSEA plot)**
- **Relative Betweenness Centrality (RBC) plot**
- **Network graph**
- **Pathway heatmap**
- **Pathway membership plot**
- **Interaction network plot**

## Installation

You can install enrichmet as:

``` r
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("enrichmet")

```

## Example

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300,
  dev = "ragg_png"  # higher quality than base png
)
```
This is a basic example 

```{r example}
## ** Example showing ALL 8 plots
# Load required libraries
library(enrichmet)
library(dplyr)
library(stringr)
library(tidyr)

## Generate example data with proper KEGG IDs
set.seed(1234)

# Create 20 unique metabolites WITH KEGG IDs
inputMetabolites <- paste0("C", sprintf("%05d", 1:20))

# ---- 1. Generate 50 pathways with random metabolites assigned ----
pathway_names <- paste0("Pathway", 1:50)
PathwayVsMetabolites <- data.frame(
    Pathway = rep(pathway_names, each = 1),
    Metabolites = sapply(1:50, function(x) paste(sample(inputMetabolites, sample(5:15, 1)), collapse = ","))
)

# ---- 2. Add new pathway entries ----
new_rows <- data.frame(
    Pathway = c("Pathway101", "Pathway102", "Pathway103", "Pathway104", "Pathway105"),
    Metabolites = c(
        "C00012,C00013,C00014,C00015,C00016,C00001,C00018,C00003,C00006,C00004",
        "C00006,C00007,C00008,C00009,C00010,C00011,C00009,C00006,C00016,C00004",
        "C00024,C00025,C00026,C00027,C00028,C00029,C00030,C00026,C00005",
        "C00013,C00014,C00015,C00016,C00017,C00024,C00027,C00014",
        "C00015,C00016,C00017,C00018,C00019,C00020,C00021,C00004,C00008,C00010"
    )
)

# Combine with existing PathwayVsMetabolites
PathwayVsMetabolites <- rbind(PathwayVsMetabolites, new_rows)

# ---- 3. Generate example metabolite-level data (with KEGG IDs) ----
example_data <- data.frame(
    met_id = c(inputMetabolites, "C00021", "C00022", "C00023"),  # Add extra metabolites
    pval = c(runif(20, 0.001, 0.05), 0.0001, 0.0002, 0.0003),
    log2fc = c(rnorm(20, mean = 0, sd = 1), 2.5, 2.3, -2.1),
    mz = rnorm(23, mean = 200, sd = 50),
    rt = rnorm(23, mean = 10, sd = 3)
)

# ---- 4. Create KEGG lookup table (ESSENTIAL) ----
kegg_lookup <- data.frame(
    kegg_id = c(inputMetabolites, "C00021", "C00022", "C00023"),
    name = c(
        "Glucose", "Lactate", "Pyruvate", "Alanine", "Valine",
        "Leucine", "Isoleucine", "Proline", "Serine", "Threonine",
        "Cysteine", "Methionine", "Aspartate", "Glutamate", "Asparagine",
        "Glutamine", "Lysine", "Arginine", "Histidine", "Phenylalanine",
        "Tyrosine", "Tryptophan", "Creatine"
    )
)

# ---- 5. Create mapping_df with additional IDs ----
set.seed(42)
mapping_df <- data.frame(
    KEGG_ID = inputMetabolites,
    HMDB_ID = paste0("HMDB", stringr::str_pad(sample(10000:99999, length(inputMetabolites)), 6, pad = "0")),
    PubChem_CID = as.character(sample(10000:99999, length(inputMetabolites))),
    CHEBI_ID = paste0("CHEBI:", sample(10000:99999, length(inputMetabolites))),
    STITCH_ID = paste0("CIDs", stringr::str_pad(sample(1000:9999, length(inputMetabolites)), 8, pad = "0")),
    Compound_Name = paste("Compound", sample(LETTERS, length(inputMetabolites), replace = TRUE), 
                          sample(100:999, length(inputMetabolites), replace = TRUE))
)

# ---- 6. Create synthetic STITCH interaction data ----
stitch_ids <- mapping_df$STITCH_ID

stitch_pairs <- expand.grid(chemical1 = stitch_ids, chemical2 = stitch_ids) %>%
    dplyr::filter(chemical1 != chemical2)

set.seed(123)
stitch_df <- stitch_pairs %>%
    dplyr::slice_sample(n = 50) %>%  # Reduced to 50 for faster processing
    dplyr::mutate(
        similarity = runif(dplyr::n(), 0.6, 0.95),  # Higher similarity for better networks
        experimental = sample(200:800, dplyr::n(), replace = TRUE),
        database = sample(c(0, 300, 600, 900), dplyr::n(), replace = TRUE),
        textmining = sample(0:900, dplyr::n(), replace = TRUE),
        combined_score = similarity * 200 + experimental + database + textmining
    )

# ---- 7. Run comprehensive enrichment analysis with ALL features ----
results <- enrichmet(
    inputMetabolites = inputMetabolites,
    PathwayVsMetabolites = PathwayVsMetabolites,
    example_data = example_data,
    kegg_lookup = kegg_lookup,
    top_n = 15,
    p_value_cutoff = 0.1,  # Use a reasonable cutoff
    mapping_df = mapping_df,
    stitch_df = stitch_df,
    analysis_type = c("enrichment", "gsea", "centrality", "network",
                      "heatmap", "membership", "interaction"),
    network_top_n = 10,
    heatmap_top_n = 10,
    membership_top_n = 10,
    min_pathway_occurrence = 2,
    min_metabolite_occurrence = 2
)
results


```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300,
  dev = "png"  # or try dev = "svg" for vector graphics
)