---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# enrichmet

<!-- badges: start -->
<!-- badges: end -->

**enrichmet** streamlines pathway enrichment analysis by enabling the entire workflow through a single R function call. This approach reduces repetitive tasks such as data formatting and parameter setup, helping users save time, minimize errors, and achieve more reproducible results.

**enrichmet** performs pathway enrichment analysis using Fisher’s exact test, computes betweenness centrality for metabolites, and performs Metabolite Set Enrichment Analysis (MetSEA). The **enrichmet**() function produces three tables (S3 data.frame objects), which may include the MetSEA table, metabolite centrality, and pathway enrichment results. In addition, it generates eight plots (S3/S4 plot objects).

## Installation

You can install enrichmet as:

``` r
install.packages("remotes")
remotes::install_git("https://github.com/biodatalab/enrichmet.git")
or 
install.packages("devtools")
devtools::install_git("https://github.com/biodatalab/enrichmet.git")
or
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("enrichmet")

```

## Example

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300,
  dev = "ragg_png"  # higher quality than base png
)
```
This is a basic example 

```{r example}
library(enrichmet)
library(dplyr)
library(stringr)

## basic example code
## ** Examples

# Generate example data with at least n=50 metabolites
set.seed(1234)

# Create 20 unique metabolites
inputMetabolites <- paste0("M", 1:20)

# ---- 1. Generate 50 pathways with random metabolites assigned ----
pathway_names <- paste0("Pathway", 1:50)
PathwayVsMetabolites <- data.frame(
  Pathway = rep(pathway_names, each = 1),
  Metabolites = sapply(1:50, function(x) paste(sample(inputMetabolites, sample(5:15, 1)), collapse = ","))
)

# ---- 2. Add new pathway entries ----
new_rows <- data.frame(
  Pathway = c("Pathway101", "Pathway102", "Pathway103", "Pathway104", "Pathway105"),
  Metabolites = c(
    "M12,M13,M14,M15,M16,M1,M18,M3,M6,M16,M4",
    "M6,M7,M8,M9,M10,M11,M9,M6,M16,M4",
    "M24,M25,M26,M27,M28,M29,M30,M26,M5",
    "M13,M14,M15,M16,M17,M24,M27,M14",
    "M15,M16,M17,M18,M19,M20,M21,M4,M8,M10"
  )
)

# Combine with existing PathwayVsMetabolites
PathwayVsMetabolites <- rbind(PathwayVsMetabolites, new_rows)

# ---- 3. Generate example metabolite-level data ----
example_data <- data.frame(
  met_id = inputMetabolites,
  pval = runif(20, 0.001, 0.05),  # Random p-values between 0.001 and 0.05
  log2fc = rnorm(20, mean = 0, sd = 1)  # Log2 fold changes from normal distribution
)

# ---- 4. Create mapping_df ----
set.seed(42)
mapping_df <- data.frame(
  KEGG_ID = inputMetabolites,
  PubChem_CID = as.character(sample(10000:99999, length(inputMetabolites))),
  STITCH_ID = paste0("CIDs", stringr::str_pad(sample(1000:9999, length(inputMetabolites)), 8, pad = "0"))
)

# ---- 5. Create synthetic STITCH interaction data ----
stitch_ids <- mapping_df$STITCH_ID

stitch_pairs <- expand.grid(chemical1 = stitch_ids, chemical2 = stitch_ids) %>%
  dplyr::filter(chemical1 != chemical2)

set.seed(123)
stitch_df <- stitch_pairs %>%
  dplyr::slice_sample(n = 200) %>%
  dplyr::mutate(
    similarity = runif(dplyr::n(), 0, 1),
    experimental = sample(0:500, dplyr::n(), replace = TRUE),
    database = sample(c(0, 300, 600, 900), dplyr::n(), replace = TRUE),
    textmining = sample(0:1000, dplyr::n(), replace = TRUE),
    combined_score = similarity * 200 + experimental + database + textmining
  )

# ---- 6. Run enrichment analysis ----
results <- enrichmet(
  inputMetabolites = inputMetabolites,
  PathwayVsMetabolites = PathwayVsMetabolites, 
  example_data = example_data,
  top_n = 10,
  mapping_df = mapping_df,
  stitch_df = stitch_df
)

# Display the results
print(results)

```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300,
  dev = "png"  # or try dev = "svg" for vector graphics
)
