---
title: "EnrichMet Tutorial"
author: "Yonatan Ayalew Mekonnen"
date: "`r Sys.Date()`"
package: "`r packageVersion('enrichmet')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: united
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Quick and Easy Pathway and Network Analysis in Metabolomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# EnrichMet

## Introduction

Advances in metabolomics have enhanced our understanding of cellular processes by enabling the detection of hundreds of metabolites in a single experiment, offering insights into complex metabolic networks. Pathway Enrichment Analysis (PEA) is key to linking these metabolites to biological pathways, but existing tools often require multiple steps, lack robust visualization, depend on web-based interfaces, or rely on difficult to install R packages—hindering reproducibility and efficiency. To address these challenges, we present EnrichMet, an R package for streamlined, reproducible pathway enrichment, metabolite set enrichment analysis (MetSEA), and network analysis. EnrichMet integrates fgsea for fast MetSEA, igraph for topology-based metrics, and curated KEGG data for enrichment using Fisher’s Exact Test—all accessible via a single function call. Through a case study on experimental metabolomics data, we show that EnrichMet outperforms existing tools in accuracy, speed, and usability. Its flexible framework also supports diverse data types and applications beyond KEGG, empowering researchers to extract deeper biological insights with minimal effort.


**EnrichMet** generates a rich set of visualizations, including:

- **Pathway enrichment plot**
- **Pathway impact plot**
- **MetSEA plot**
- **RBC plot**
- **Network graph**
- **Pathway heatmap**
- **Pathway membership plot**
- **Interaction network plot**

Additionally, EnrichMet exports both the **MetSEA** and **pathway enrichment** results as `.xlsx` tables, streamlining both analysis and reporting.


## Installation

You can install EnrichMet as:

```r
install.packages("remotes")
remotes::install_git("https://github.com/biodatalab/enrichmet.git")
# or 
install.packages("devtools")
devtools::install_git("https://github.com/biodatalab/enrichmet.git")
# or
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("enrichmet")

```

### Dependancies

```{r, Case study of KrasG12D mutation}
library(igraph)
library(ggraph)
library(ggplot2)
library(openxlsx)
library(dplyr)
library(tidyr)
library(fgsea)
library(ggrepel)
library(stringr)
library(ComplexHeatmap)
library(readr)
library(magrittr)
library(circlize)
library(scales)
library(tibble)
library(knitr)
library(httr)
```
## Required Input Files

To run the **EnrichMet** package, you need to provide at least the following three files:

- **`PathwayVsMetabolites`**  
  This file contains a table mapping metabolic pathways to their associated metabolites, defining which metabolites belong to each pathway for enrichment analysis.  
  We provide a human-specific `PathwayVsMetabolites` file available here:  
  [Human-PathwaysVsMetabolites.csv](https://zenodo.org/api/records/15498097/files/Human-PathwaysVsMetabolites.csv)  
  However, users can supply a file tailored to their species of interest.

- **`InputMetabolites`**  
  A list of metabolites from your experimental data to analyze, typically consisting of metabolite identifiers detected or quantified in your study. This input file should be prepared and provided by the user.

- **`SummaryStatistics`**  
  Statistical results corresponding to the input metabolites, such as fold changes, p-values, or other relevant metrics. These statistics guide the enrichment analysis by reflecting the significance or effect size of each metabolite. This file is also user-provided.

## Optional Input Files

Including these optional files can enhance the analysis and its biological interpretation:

- **`KEGGLookup`**  
  A reference file linking KEGG compound IDs to metabolite names or other annotations, which helps standardize identifiers and improve pathway mapping.  
  We provide a KEGGLookup file here:  
  [kegg_lookup.xlsx](https://zenodo.org/api/records/15498097/files/kegg_lookup.xlsx)  
  Users may also provide a lookup file specific to their species.

- **`STITCHInteractions`**  
  A dataset of known or predicted interactions between metabolites or proteins, useful for constructing interaction networks to support pathway analysis.  
  We supply an example file here:  
  [chemical_chemical.tsv](https://zenodo.org/api/records/15498097/files/chemical_chemical.tsv)  
  Users may provide their own interaction data appropriate for their species.

- **`mapping_df`**  
  A mapping data frame that links STITCH identifiers with metabolite names, facilitating integration between datasets.  
  The provided example file is available here:  
  [mapping_df.xlsx](https://zenodo.org/api/records/15498097/files/mapping_df.xlsx)  
  Users can substitute this with a mapping file relevant to their species of interest.


## Case study of KrasG12D mutation

We used summary statistics from differential analyses based on metabolomic measurements comparing KrasG12D mutations with and without TAp73 deletion. The top 50 metabolites, ranked by significance, were selected for pathway enrichment analysis using Fisher’s Exact Test. These summary statistics were also utilized for MetSEA analysis.

```{r}
# InputMetabolites: A vector or list containing the metabolites of interest.
inputMetabolites <- c(
  "C02721", "C03736", "C03139", "C00074", "C02862", "C00092", "C00275",
  "C00386", "C10172", "C00361", "C00242", "C19806", "C16358", "C16357",
  "C16353", "C02140", "C05488", "C01181", "C03017", "C02632", "C00246",
  "C00079", "C08317", "C00588", "C16527", "C00065", "C00249", "C00836",
  "C02990", "C03690", "C04025", "C06424", "C00123", "C00606", "C03067",
  "C00539", "C00633", "C00519", "C00366", "C00148", "C00219", "C08261",
  "C00385", "C06429", "C05637", "C00295", "C06428", "C16513", "C01909",
  "C03299", "C01530", "C01571", "C08278", "C00599", "C15587", "C00144",
  "C03739"
)
# PathwayVsMetabolites: A file mapping pathways to their corresponding metabolites.
PathwayVsMetabolites=read.csv("https://zenodo.org/api/records/15498097/files/Human-PathwaysVsMetabolites.csv/content")
str(PathwayVsMetabolites)
# SummaryStatistics: A file containing differential analysis results or other relevant statistical outputs.
example_data=read.xlsx("https://zenodo.org/api/records/15498097/files/example_data.xlsx/content")
head(example_data)
# KEGGLookup: A file containing annotated KEGG compound IDs with their corresponding metabolite names.
kegg_lookup=read.xlsx("https://zenodo.org/api/records/15498097/files/kegg_lookup.xlsx/content")
head(kegg_lookup)
# Mapping_df: A data frame that maps metabolites to their corresponding STITCH IDs.
mapping_df = read.xlsx("https://zenodo.org/api/records/15498097/files/mapping_df.xlsx/content")
head(mapping_df)
#STITCHInteractions: A file obtained from the STITCH database containing chemical–chemical interaction information.
options(timeout = 600)  
url <- "https://zenodo.org/api/records/15498097/files/chemical_chemical.tsv/content"
response <- GET(url, config = config(ssl_verifypeer = FALSE))
stitch_df <- read_tsv(content(response, "raw"))
head(stitch_df)

#  Run enrichment analysis 
#loading the package
library(enrichmet)
results <-enrichmet(
  inputMetabolites = inputMetabolites,
  PathwayVsMetabolites = PathwayVsMetabolites,example_data, kegg_lookup,
  top_n = 100,p_value_cutoff = 1,
  mapping_df = mapping_df,
  stitch_df = stitch_df
)

```

## Display Plots

```{r display-plots1, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$pathway_plot)
```
The plot displays the enriched pathways along with their corresponding -log10 p-values. Metabolite lists were used as input, and pathway enrichment was performed using Fisher's exact test based on metabolite-to-pathway mapping.

```{r display-plots2, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$impact_plot)
```

The plot presents the pathway impact analysis, combining enrichment significance (p-value) with Relative Betweenness Centrality (RBC) to highlight both statistical and topological relevance.

```{r display-plots3, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$MetSEA_plot)
```
The plot displays the Metabolite Set Enrichment Analysis (MetSEA) results, highlighting significantly enriched metabolite sets

```{r display-plots4, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$rbc_plot)
```
The plot illustrates the relative betweenness centrality of the input metabolites, indicating their topological importance and potential regulatory influence within the metabolic network.
```{r display-plots5, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$network_plot)
```
The plot presents the network graph, highlighting the relationships between enriched pathways and input metabolites, and illustrating their connectivity and functional associations within the metabolic network.
```{r display-plots6, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$heatmap_plot)
```
The heatmap displays the gradient of -log p-values, indicating the significance levels of the enriched pathways.

```{r display-plots7, results='asis', fig.align='center', fig.width=15, fig.height=10}
cat('<div style="overflow-x: auto; width: 100%;">')
ComplexHeatmap::draw(results$membership_plot)
cat('</div>')
```
The plot illustrates pathway membership by showing each pathway along with its associated metabolites.
```{r display-plots8, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$interaction_plot)
```
The plot displays metabolite interactions derived from chemical-chemical interaction data obtained from STITCH.

```{r session-info, echo=FALSE}
sessionInfo()
```

