---
title: "enrichmet Tutorial"
author: "Yonatan Ayalew Mekonnen"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{1. Quick and Easy Pathway and Network Analysis in Metabolomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Introduction

Advances in metabolomics have enhanced our understanding of cellular processes by enabling the detection of hundreds of metabolites in a single experiment, offering insights into complex metabolic networks. Pathway Enrichment Analysis (PEA) is key to linking these metabolites to biological pathways, but existing tools often require multiple steps, lack robust visualization, depend on web-based interfaces, or rely on difficult to install R packages—hindering reproducibility and efficiency. To address these challenges, we present **enrichmet**, an R package for streamlined, reproducible pathway enrichment, metabolite set enrichment analysis (MetSEA), and network analysis. **enrichmet** integrates fgsea for fast MetSEA, igraph for topology-based metrics, and curated KEGG data for enrichment using Fisher’s Exact Test—all accessible via a single function call. Through a case study on experimental metabolomics data, we show that **enrichmet** outperforms existing tools in accuracy, speed, and usability. Its flexible framework also supports diverse data types and applications beyond KEGG, empowering researchers to extract deeper biological insights with minimal effort.

**enrichmet** streamlines pathway enrichment analysis by enabling the entire workflow through a single R function call. This approach reduces repetitive tasks such as data formatting and parameter setup, helping users save time, minimize errors, and achieve more reproducible results. For added flexibility, the package also offers modular functions that allow users to tailor the workflow to specific needs. **enrichmet** emphasizes transparency and control, making outputs easy to access and integrate into broader R workflows. Its versatile framework extends beyond metabolomics, with applicability to lipidomics, proteomics, and other omics domains. Importantly, because enrichment is performed locally—without dependence on external web services, **enrichmet** is especially well suited for high-throughput and iterative analyses where speed and reproducibility are critical.

**enrichmet** generates a rich set of visualizations, including:

- **Pathway enrichment plot**
- **Pathway impact plot**
- **MetSEA plot**
- **RBC plot**
- **Network graph**
- **Pathway heatmap**
- **Pathway membership plot**
- **Interaction network plot**

The enrichmet() function produces three tables (S3 data.frame objects), which may optionally include the MSEA table, metabolite centrality, and pathway enrichment results. In addition, it generates eight plots (S3/S4 plot objects) as described above.

# Installation

You can install enrichmet as:

```r
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("enrichmet")

```

```{r, Case study of KrasG12D mutation}
library(BiocFileCache)
library(readxl)
library(readr)
library(enrichmet)
```
## Required Input Files

To run the **enrichmet** package, you need to provide at least the following three files:

- **`PathwayVsMetabolites`**  
This file defines the mapping between metabolic pathways and their associated metabolites, serving as the background reference for Fisher’s exact test–based enrichment analysis (see section 2.4).

The mappings are derived from the KEGG database (https://www.kegg.jp/), accessed programmatically via the KEGGREST package. To ensure reproducibility and minimize run-time delays, curated human-specific pathway–metabolite mappings are periodically updated and archived on Zenodo [1].
  We provide a human-specific `PathwayVsMetabolites` file available here:  
  [Human-PathwaysVsMetabolites.csv](https://zenodo.org/api/records/15498097/files/Human-PathwaysVsMetabolites.csv)  
  However, users can supply a file tailored to their species of interest.

- **`InputMetabolites`**  
A list of metabolites from the user’s experimental dataset for analysis, typically consisting of metabolite identifiers (e.g., significant metabolites from differential analysis (see section 2.4)). This input is not provided by the package and must be prepared by the user. The file serves as the query set for enrichment analysis, and its identifiers should match those used in the background pathway–metabolite mappings (e.g., KEGG compound IDs).

For demonstration purposes, we also include summary statistics derived from an example experimental dataset. These example data are intended solely to showcase the package workflow and will be described in detail in a subsequent publication. 


- **`SummaryStatistics`**  
The enrichment analysis requires statistical results associated with the input metabolites, such as fold changes, p-values, or other relevant effect size metrics (see section 2.4). These values guide the enrichment procedure by reflecting both the significance and the magnitude of change for each metabolite.

This information is not generated by the package and must be provided by the user. Summary statistics may be supplied either as a standalone data frame or extracted directly from the rowData() of a SummarizedExperiment object, provided that appropriate columns are available for metabolite identifiers, p-values, and fold changes (see section 2.4 example_data).

For demonstration purposes, the package includes an example summary statistics dataset. These example data are provided solely to illustrate the workflow and will be described in detail in a forthcoming publication.

The **enrichmet**() function integrates both p-values and fold changes when performing MetSEA. By considering the direction and magnitude of changes, this approach yields richer biological insights than methods based only on binary classifications of "significant" versus "not significant."

## Optional Input Files

Including these optional files can enhance the analysis and its biological interpretation:

- **`KEGGLookup`**  
  A reference file linking KEGG compound IDs (https://www.kegg.jp/) to metabolite names or other annotations, which helps standardize identifiers and improve pathway mapping [1].
  We provide a KEGGLookup file here:  
  [kegg_lookup.xlsx](https://zenodo.org/api/records/15498097/files/kegg_lookup.xlsx)  
  Users may also provide a lookup file specific to their species.

- **`STITCHInteractions`**  
A dataset of known or predicted interactions between metabolites or proteins, obtained from the freely available STITCH database (http://stitch.embl.de/
). These interactions are useful for constructing networks to support pathway analysis [2].
  We supply an example file here:  
  [chemical_chemical.tsv](https://zenodo.org/api/records/15498097/files/chemical_chemical.tsv)  
  Users may provide their own interaction data appropriate for their species.

- **`mapping_df`**  
  A mapping data frame that links STITCH identifiers with metabolite names, facilitating integration between datasets [2].
  The provided example file is available here:  
  [mapping_df.xlsx](https://zenodo.org/api/records/15498097/files/mapping_df.xlsx)  
  Users can substitute this with a mapping file relevant to their species of interest.


## Case study of KrasG12D mutation

We used summary statistics from differential analyses based on metabolomic measurements comparing KrasG12D mutations with and without TAp73 deletion. The top 50 metabolites, ranked by significance, were selected for pathway enrichment analysis using Fisher’s Exact Test. These summary statistics were also utilized for MetSEA analysis.

```{r}
# InputMetabolites: A vector or list containing the metabolites of interest.
inputMetabolites <- c(
  "C02721", "C03736", "C03139", "C00074", "C02862", "C00092", "C00275",
  "C00386", "C10172", "C00361", "C00242", "C19806", "C16358", "C16357",
  "C16353", "C02140", "C05488", "C01181", "C03017", "C02632", "C00246",
  "C00079", "C08317", "C00588", "C16527", "C00065", "C00249", "C00836",
  "C02990", "C03690", "C04025", "C06424", "C00123", "C00606", "C03067",
  "C00539", "C00633", "C00519", "C00366", "C00148", "C00219", "C08261",
  "C00385", "C06429", "C05637", "C00295", "C06428", "C16513", "C01909",
  "C03299", "C01530", "C01571", "C08278", "C00599", "C15587", "C00144",
  "C03739"
)
# Standard cache location
cache_dir <- tools::R_user_dir("enrichmet", "cache")
bfc <- BiocFileCache(cache_dir, ask = FALSE)

# Helper function to download & cache any file from Zenodo
get_cached_file <- function(url) {
  bfcrpath(bfc, url)
}

# -------------------------
# Load files with caching
# -------------------------

# PathwayVsMetabolites
pathway_url <- "https://zenodo.org/api/records/15498097/files/Human-PathwaysVsMetabolites.csv/content"
pathway_path <- get_cached_file(pathway_url)
PathwayVsMetabolites <- read.csv(pathway_path)
str(PathwayVsMetabolites)

# SummaryStatistics
example_url <- "https://zenodo.org/api/records/15498097/files/example_data.xlsx/content"
example_path <- get_cached_file(example_url)
example_data <- read_excel(example_path)
head(example_data)

# KEGGLookup
kegg_url <- "https://zenodo.org/api/records/15498097/files/kegg_lookup.xlsx/content"
kegg_path <- get_cached_file(kegg_url)
kegg_lookup <- read_excel(kegg_path)
head(kegg_lookup)

# Mapping_df
mapping_url <- "https://zenodo.org/api/records/15498097/files/mapping_df.xlsx/content"
mapping_path <- get_cached_file(mapping_url)
mapping_df <- read_excel(mapping_path)
head(mapping_df)

# STITCHInteractions
stitch_url <- "https://zenodo.org/records/15498097/files/chemical_chemical.tsv"
stitch_path <- get_cached_file(stitch_url)
stitch_df <- read_tsv(stitch_path)
head(stitch_df)
#  Run enrichment analysis 
#loading the package
results <-enrichmet(
  inputMetabolites = inputMetabolites,
  PathwayVsMetabolites = PathwayVsMetabolites,example_data, kegg_lookup,
  top_n = 100,p_value_cutoff = 1,
  mapping_df = mapping_df,
  stitch_df = stitch_df
)

```

## Display Plots

```{r display-plots1, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$pathway_plot)
```
Fig. 1. Pathway enrichment plot. The plot shows enriched pathways with their corresponding –log10 p-values. Circle size represents the number of metabolites associated with each pathway, while color indicates the adjusted p-value. Pathway enrichment was performed using Fisher’s exact test based on metabolite-to-pathway mappings, with metabolite lists as input.

```{r display-plots2, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$impact_plot)
```

Fig. 2. Pathway impact plot. The plot integrates enrichment significance with pathway topology to highlight the most influential pathways. The y-axis shows the −log10 p-value and x-axis shows the pathway impact value. Circle size reflects the pathway impact, while color represents the adjusted p-value. Together, these features emphasize pathways that are both statistically significant and topologically central.

```{r display-plots3, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$gsea_plot)
```
Fig. 3. Metabolite Set Enrichment Analysis (MetSEA) plot. The plot shows enriched pathways with their corresponding –log10 p-values. Circle size represents the number of metabolites associated with each pathway, while color indicates the Normalized Enrichment Score (NES).

```{r display-plots4, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$rbc_plot)
```
Fig. 4. Relative Betweenness Plot. This plot displays the Relative Betweenness Centrality (RBC) of input metabolites, with RBC on the x-axis and corresponding values on the y-axis. It highlights the topological importance and potential regulatory influence of metabolites within the metabolic network.
```{r display-plots5, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$network_plot)
```
Fig. 5. Pathway–Metabolite Network Plot. This plot visualizes the relationships between enriched pathways and input metabolites, highlighting their connectivity and functional associations within the metabolic network. Edges indicate associations between pathways and their corresponding metabolites, while nodes are colored to distinguish pathways from metabolites.
```{r display-plots6, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$heatmap_plot)
```
Fig. 6. Pathway Heatmap. Metabolites are displayed on the y-axis and pathways on the x-axis. The heatmap uses a gradient of −log10 p-values to indicate the significance of enriched pathways and the involvement of each metabolite within the corresponding pathway.

```{r display-plots7, results='asis', fig.align='center', fig.width=15, fig.height=10}

ComplexHeatmap::draw(results$membership_plot)

```
Fig. 7. Pathway Membership plot. Metabolites are displayed on the y-axis and pathways on the x-axis. The heatmap plot illustrates pathway membership by showing each pathway along with its associated metabolites.Only pathways with at least two mapped metabolites are included.
```{r display-plots8, results='asis', fig.align='center', fig.width=12, fig.height=10}
print(results$interaction_plot)
```
Fig. 8. Chemical-Chemical Interaction plot. The plot displays metabolite interactions derived from chemical-chemical interaction data obtained from STITCH. Metabolites are shown as nodes whose size scales with degree (the number of direct interaction partners). Edges represent experimentally validated or predicted biochemical relationship; and  edge opacity is mapped to the rescaled STITCH combined score (weight), with darker edges indicating higher-confidence interaction. 

## References

1. Kanehisa, M., Furumichi, M., Sato, Y., Matsuura, Y., & Ishiguro-Watanabe, M. (2025). KEGG: biological systems database as a model of the real world. *Nucleic Acids Research*, 53, D672–D677. [https://www.kegg.jp/](https://www.kegg.jp/)

2. Szklarczyk, D., Santos, A., von Mering, C., Jensen, L. J., Bork, P., & Kuhn, M. (2016). STITCH 5: augmenting protein–chemical interaction networks with tissue and affinity data. *Nucleic Acids Research*, 44(D1), D380–D384. [https://stitch.embl.de/](https://stitch.embl.de/)



```{r session-info, echo=FALSE}
# reproducibility
sessionInfo()
```

