---
title: "enrichmet Tutorial"
author: "Yonatan Ayalew Mekonnen"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{1. Quick and Easy Pathway and Network Analysis in Metabolomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Introduction

Advances in metabolomics have enhanced our understanding of cellular processes by enabling the detection of hundreds of metabolites in a single experiment, offering insights into complex metabolic networks. Pathway Enrichment Analysis (PEA) is key to linking these metabolites to biological pathways, but existing tools often require multiple steps, lack robust visualization, depend on web-based interfaces, or rely on difficult to install R packages—hindering reproducibility and efficiency. To address these challenges, we present enrichmet, an R package for streamlined, reproducible pathway enrichment, metabolite set enrichment analysis (MetSEA), and network analysis. enrichmet integrates fgsea for fast MetSEA, igraph for topology-based metrics, and curated KEGG data for enrichment using Fisher’s Exact Test—all accessible via a single function call. Through a case study on experimental metabolomics data, we show that enrichmet outperforms existing tools in accuracy, speed, and usability. Its flexible framework also supports diverse data types and applications beyond KEGG, empowering researchers to extract deeper biological insights with minimal effort.


**enrichmet** generates a rich set of visualizations, including:

- **Pathway enrichment plot**
- **Pathway impact plot**
- **MetSEA plot**
- **RBC plot**
- **Network graph**
- **Pathway heatmap**
- **Pathway membership plot**
- **Interaction network plot**

Additionally, enrichmet exports both the **MetSEA** and **pathway enrichment** results as `.xlsx` tables, streamlining both analysis and reporting.


# Installation

You can install enrichmet as:

```r
install.packages("remotes")
remotes::install_git("https://github.com/biodatalab/enrichmet.git")
# or 
install.packages("devtools")
devtools::install_git("https://github.com/biodatalab/enrichmet.git")
# or
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("enrichmet")

```

## Dependancies

```{r, Case study of KrasG12D mutation}
library(igraph)
library(ggraph)
library(ggplot2)
library(readxl)
library(dplyr)
library(tidyr)
library(fgsea)
library(ggrepel)
library(stringr)
library(ComplexHeatmap)
library(readr)
library(magrittr)
library(circlize)
library(scales)
library(tibble)
library(knitr)
library(httr)
library(BiocFileCache)
```
## Required Input Files

To run the **enrichmet** package, you need to provide at least the following three files:

- **`PathwayVsMetabolites`**  
This file defines the mapping between metabolic pathways and their associated metabolites, serving as the background reference for Fisher’s exact test–based enrichment analysis.

The mappings are derived from the KEGG database (https://www.kegg.jp/), accessed programmatically via the KEGGREST package. To ensure reproducibility and minimize run-time delays, curated human-specific pathway–metabolite mappings are periodically updated and archived on Zenodo [1].
  We provide a human-specific `PathwayVsMetabolites` file available here:  
  [Human-PathwaysVsMetabolites.csv](https://zenodo.org/api/records/15498097/files/Human-PathwaysVsMetabolites.csv)  
  However, users can supply a file tailored to their species of interest.

- **`InputMetabolites`**  
A list of metabolites from the user’s experimental dataset to be analyzed, typically consisting of metabolite identifiers that were detected or quantified in the study. This input is not provided by the package and must be prepared by the user. The file serves as the query set for enrichment analysis, and its identifiers should match those used in the background pathway–metabolite mappings (e.g., KEGG compound IDs).

For demonstration purposes, we also include summary statistics derived from an example experimental dataset. These example data are intended solely to showcase the package workflow and will be described in detail in a subsequent publication.

- **`SummaryStatistics`**  
Statistical results corresponding to the input metabolites, such as fold changes, p-values, or other relevant metrics. These results guide the enrichment analysis by reflecting the significance or effect size of each metabolite.

This file is not provided by the package and must be prepared by the user. The information may be supplied as a separate data frame or retrieved directly from the rowData() of a SummarizedExperiment object, provided that it contains appropriate columns for metabolite identifiers, p-values, and fold changes.

For demonstration purposes, the package includes an example summary statistics dataset. These example data are included solely to illustrate the workflow and will be fully described in a subsequent publication.

## Optional Input Files

Including these optional files can enhance the analysis and its biological interpretation:

- **`KEGGLookup`**  
  A reference file linking KEGG compound IDs (https://www.kegg.jp/) to metabolite names or other annotations, which helps standardize identifiers and improve pathway mapping [1].
  We provide a KEGGLookup file here:  
  [kegg_lookup.xlsx](https://zenodo.org/api/records/15498097/files/kegg_lookup.xlsx)  
  Users may also provide a lookup file specific to their species.

- **`STITCHInteractions`**  
A dataset of known or predicted interactions between metabolites or proteins, obtained from the freely available STITCH database (http://stitch.embl.de/
). These interactions are useful for constructing networks to support pathway analysis [2].
  We supply an example file here:  
  [chemical_chemical.tsv](https://zenodo.org/api/records/15498097/files/chemical_chemical.tsv)  
  Users may provide their own interaction data appropriate for their species.

- **`mapping_df`**  
  A mapping data frame that links STITCH identifiers with metabolite names, facilitating integration between datasets [2].
  The provided example file is available here:  
  [mapping_df.xlsx](https://zenodo.org/api/records/15498097/files/mapping_df.xlsx)  
  Users can substitute this with a mapping file relevant to their species of interest.


## Case study of KrasG12D mutation

We used summary statistics from differential analyses based on metabolomic measurements comparing KrasG12D mutations with and without TAp73 deletion. The top 50 metabolites, ranked by significance, were selected for pathway enrichment analysis using Fisher’s Exact Test. These summary statistics were also utilized for MetSEA analysis.

```{r}
# InputMetabolites: A vector or list containing the metabolites of interest.
inputMetabolites <- c(
  "C02721", "C03736", "C03139", "C00074", "C02862", "C00092", "C00275",
  "C00386", "C10172", "C00361", "C00242", "C19806", "C16358", "C16357",
  "C16353", "C02140", "C05488", "C01181", "C03017", "C02632", "C00246",
  "C00079", "C08317", "C00588", "C16527", "C00065", "C00249", "C00836",
  "C02990", "C03690", "C04025", "C06424", "C00123", "C00606", "C03067",
  "C00539", "C00633", "C00519", "C00366", "C00148", "C00219", "C08261",
  "C00385", "C06429", "C05637", "C00295", "C06428", "C16513", "C01909",
  "C03299", "C01530", "C01571", "C08278", "C00599", "C15587", "C00144",
  "C03739"
)
# Standard cache location
cache_dir <- tools::R_user_dir("enrichmet", "cache")
bfc <- BiocFileCache(cache_dir, ask = FALSE)

# Helper function to download & cache any file from Zenodo
get_cached_file <- function(url) {
  bfcrpath(bfc, url)
}

# -------------------------
# Load files with caching
# -------------------------

# PathwayVsMetabolites
pathway_url <- "https://zenodo.org/api/records/15498097/files/Human-PathwaysVsMetabolites.csv/content"
pathway_path <- get_cached_file(pathway_url)
PathwayVsMetabolites <- read.csv(pathway_path)
str(PathwayVsMetabolites)

# SummaryStatistics
example_url <- "https://zenodo.org/api/records/15498097/files/example_data.xlsx/content"
example_path <- get_cached_file(example_url)
example_data <- read_excel(example_path)
head(example_data)

# KEGGLookup
kegg_url <- "https://zenodo.org/api/records/15498097/files/kegg_lookup.xlsx/content"
kegg_path <- get_cached_file(kegg_url)
kegg_lookup <- read_excel(kegg_path)
head(kegg_lookup)

# Mapping_df
mapping_url <- "https://zenodo.org/api/records/15498097/files/mapping_df.xlsx/content"
mapping_path <- get_cached_file(mapping_url)
mapping_df <- read_excel(mapping_path)
head(mapping_df)

# STITCHInteractions
stitch_url <- "https://zenodo.org/records/15498097/files/chemical_chemical.tsv"
stitch_path <- get_cached_file(stitch_url)
stitch_df <- read_tsv(stitch_path)
head(stitch_df)
#  Run enrichment analysis 
#loading the package
library(enrichmet)
results <-enrichmet(
  inputMetabolites = inputMetabolites,
  PathwayVsMetabolites = PathwayVsMetabolites,example_data, kegg_lookup,
  top_n = 100,p_value_cutoff = 1,
  mapping_df = mapping_df,
  stitch_df = stitch_df
)

```

## Display Plots

```{r display-plots1, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$pathway_plot)
```
The plot displays the enriched pathways along with their corresponding -log10 p-values. Metabolite lists were used as input, and pathway enrichment was performed using Fisher's exact test based on metabolite-to-pathway mapping.

```{r display-plots2, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$impact_plot)
```

The plot presents the pathway impact analysis, combining enrichment significance (p-value) with Relative Betweenness Centrality (RBC) to highlight both statistical and topological relevance.

```{r display-plots3, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$MetSEA_plot)
```
The plot displays the Metabolite Set Enrichment Analysis (MetSEA) results, highlighting significantly enriched metabolite sets

```{r display-plots4, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$rbc_plot)
```
The plot illustrates the relative betweenness centrality of the input metabolites, indicating their topological importance and potential regulatory influence within the metabolic network.
```{r display-plots5, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$network_plot)
```
The plot presents the network graph, highlighting the relationships between enriched pathways and input metabolites, and illustrating their connectivity and functional associations within the metabolic network.
```{r display-plots6, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$heatmap_plot)
```
The heatmap displays the gradient of -log p-values, indicating the significance levels of the enriched pathways.

```{r display-plots7, results='asis', fig.align='center', fig.width=15, fig.height=10}
cat('<div style="overflow-x: auto; width: 100%;">')
ComplexHeatmap::draw(results$membership_plot)
cat('</div>')
```
The plot illustrates pathway membership by showing each pathway along with its associated metabolites.
```{r display-plots8, results='asis', fig.align='center', fig.width=12, fig.height=8}
print(results$interaction_plot)
```
The plot displays metabolite interactions derived from chemical-chemical interaction data obtained from STITCH.

## References

1. Kanehisa, M., Furumichi, M., Sato, Y., Matsuura, Y., & Ishiguro-Watanabe, M. (2025). KEGG: biological systems database as a model of the real world. *Nucleic Acids Research*, 53, D672–D677. [https://www.kegg.jp/](https://www.kegg.jp/)

2. Szklarczyk, D., Santos, A., von Mering, C., Jensen, L. J., Bork, P., & Kuhn, M. (2016). STITCH 5: augmenting protein–chemical interaction networks with tissue and affinity data. *Nucleic Acids Research*, 44(D1), D380–D384. [https://stitch.embl.de/](https://stitch.embl.de/)



```{r session-info, echo=FALSE}
# reproducibility
sessionInfo()
```

